---
title: "Examples of querying GLORYS datas"
format: html
editor: visual
---

## Downloading – straight Python

This is the approach I'm using the most right now – querying temp file downloads straight from Python. You have to install via <https://pypi.org/project/copernicusmarine/> – I like this because I think it's made by the Copernicus development team and seems very stable.

```{python echo=FALSE}
# example of 
copernicusmarine.subset(
  dataset_id="cmems_mod_glo_phy_my_0.083deg_P1M-m",
  variables=["thetao"],  # temperature
  minimum_longitude=-126,
  maximum_longitude=-115,
  minimum_latitude=32,
  maximum_latitude=50,
  start_datetime="1994-01-01T00:00:00",
  end_datetime="2024-08-20T23:59:59",
  output_directory="./copernicus-data",  # Directory to save 
  output_filename="glorys_data_01.nc"
)

```

## Downloading – reticulate

As an alternative to pure Python, you can use the reticulate package in R.

```{r}
library(reticulate)
#install_python() 
virtualenv_create(envname = "thiscopernicus")
#virtualenv_install("thiscopernicus", packages = c("copernicusmarine"))
use_virtualenv("thiscopernicus", required = TRUE)
#py_install("copernicusmarine")

cm <- import("copernicusmarine")

cm$subset(
    dataset_id = "cmems_mod_glo_phy_my_0.083deg_P1M-m",
    start_datetime=paste0(1994,"-01-01T00:00:00"),
    end_datetime=paste0(2024,"-12-31T23:59:59"),
    variables = list("thetao"),
    minimum_longitude = -126,
    maximum_longitude = -115,
    minimum_latitude = 32,
    maximum_latitude = 50,
    minimum_depth = 50,
    maximum_depth = 250,
    output_filename = "CMEMS_tmp.nc"
  )
```

## Downloading – CopernicusMarine

This example works with the R package CopernicusMarine. You need to authenticate credentials the first time you use it, and I've avoided it because it was a little glitchy during the last Copernicus upgrade

```{r}
library(CopernicusMarine)
my_data <- cms_download_subset(
  product = "GLOBAL_MULTIYEAR_PHY_001_030",
  layer = "cmems_mod_glo_phy_my_0.083deg_P1M-m",
  variable = "thetao",
  region = c(-126, 32, -115, 50), # lon_min, lat_min, lon_max, lat_max
  timerange = c("1994-01-01", "2024-08-20"),
  verticalrange = c(-50, -250), # 50m to 250m depth
  progress = TRUE
)
```

## Processing – netCDF

For most of these, you can use `tidync` in R to process the data

```{r}
  temp <- tidync::tidync("CMEMS_tmp.nc") |>
    hyper_tibble( force = TRUE) |>
    drop_na() |>
    group_by(longitude,latitude,time)
```

## Post processing – summarising into time series

```{r}
library(lubridate)
library(tidync)
library(tidyverse)
library(data.table)
library(CopernicusMarine)
```

Example time series: MLDyolk To get the appropriate time series, we need to summarise the dataset. We can easily just download the data at the depths and latitude we need but to get this time series as "between the 100 and 2,000 m isobath" as defined in Vestfals et al. 2023 and to do that we need to combine our .nc tibble with a bathymetric dataset. To approach this I am going to download beyond the longitude I need then combine datasets. You can download the bathymetry data from GLORYS and then use the grid to assign longitude. Here I am just going to use the dataset as a .csv that Nick and I are already using. No need to reinvent the wheel...

First lets download the mixed layer data, I usually use reticulate so I will just use Eric's approach from above:

```{r}

cm <- import("copernicusmarine")

cm$subset(
    dataset_id = "cmems_mod_glo_phy_my_0.083deg_P1M-m",
    start_datetime=paste0(1994,"-01-01T00:00:00"),
    end_datetime=paste0(2024,"-12-31T23:59:59"),
    variables = list("mlotst"),
    minimum_longitude = -130, #probably more longitude than we need
    maximum_longitude = -115,
    minimum_latitude = 31,
    maximum_latitude = 36,#no depth because MLD is not a variable with depths
    output_filename = "CMEMS_mlotst.nc"
  )
```

Summarise just as Eric did

```{r}

  temp <- tidync::tidync("CMEMS_mlotst.nc") |>
    hyper_tibble( force = TRUE) |>
    drop_na() |>
    group_by(longitude,latitude,time)
```

Copying in what we have been using for this - this is a static variable from GLORYS and linked to lat/long using the grid cell info

```{r}

bath <- read.csv('databathymetry-m.csv')

combined <- bath%>%left_join(temp)%>% #combining data so we have bottom depth to filter over for isobath
  filter(bottom_depth >= 100 & bottom_depth <= 2000)%>% #isobath greater than 100m less than 2,000
  mutate(date = as_date(time/24,origin="1950-01-01 00:00:00"))%>% #adding in date
  mutate(year=year(date), month = month(date))%>% #getting out year and month from date
  filter(month ==1|month==2|month==3|month==4 )%>% #getting the months for the yolk-sac lifestage
  group_by(year)%>%
  summarise(MLD_yolk_New=mean(mlotst)) #woo! we have it



```

I am just going to double check this against the old dataset that I have for hake to make sure we didnt mess anything up...you could swap out the file for whatever you have from last years assessment (I think that is what I have...)

```{r}

old <- read.csv('DATA_Combined_glorys_hake_UNSTANDARDIZED.csv')%>%
  select(year, MLDyolk)%>%
  rename(MLDyolk_Old=MLDyolk)

comparison_wide <- old%>%left_join(combined)
comparison<-  comparison_wide%>%pivot_longer(-year)

ggplot(data=comparison, aes(y=value, x=year, col=name))+
  geom_line()+
  theme_bw() #looks great! These minor differences are from the reanalysis

cor(comparison_wide%>%filter(year>1993&year<2024)%>%select(MLDyolk_Old), 
    comparison_wide%>%filter(year>1993&year<2024)%>%select(MLD_yolk_New)) #0.99 love to see it

```
