---
title: "MOM6_variables"
format: pdf
editor: source
---

# Setup

First we need to set up our python environment for reticulate and make sure it is calling the correct version of icechunk and arraylake to be able to interact with Eartmover. Reticulate automatically loads an environment when you start a new session and ensuring the python environment you are using with reticulate is correctly specifying the correct version of python and the Earthmover associated packages is key.

```{r}
library(reticulate)

# Force reticulate to pull down the newest package versions
#reticulate::py_install("dask")
py_require("icechunk>=2.0.0")
py_require("arraylake>=1.0.0")
py_require("zarr")
py_require("h5netcdf")
py_require("h5py")
py_require("xarray")
py_require("scipy")
py_require("dask")
# Import the fresh libraries
arraylake <- import("arraylake") #the data visual layer of Earthmover
zarr <- import("zarr") #interact with compressed data chunks and JSON metadata
xr   <- import("xarray") #python library for working with arrays
np   <- np <- import("numpy", convert = FALSE) #necessary for the NaN checking block
pd   <- import("pandas")
time <- import("time")
dask <- import("dask")

# Verify the version 
print(arraylake$`__version__`) 
```

Now we will load the data from Earthmover and a static .netcdf. Whether or not this loads correctly will identify whether the python environment is loaded correctly in R and may require trounblshooting 
```{r}
# Connect to your NOAA data repository
client <- arraylake$Client()
#token issue? go to terminal arraylake auth login
repo <- client$get_repo("NOAA-PMEL/cefi-nep-hindcast-monthly") 
print(repo)

#If this runs then Earthmover is functioning!

ds_depth = xr$open_dataset("../STATIC/ocean_static.deptho.nc")

#If this runs then xarray has appropriate dependencies loaded

```



```{r}
# assign months, latitudes, and depths to filter
# all are integers, depths are in meters
# filters are inclusive (closed interval), e.g. depth 0-180 includes depths 0 and 180 and every depth between

lifestage_dict <- list(
  spawn = list( # spawn
    min_month = 1,
    max_month = 3,
    min_lat   = 31,
    max_lat   = 36,
    min_depth = 130,
    max_depth = 500
  ), #long extent shelf break
  egg = list( # egg
    min_month = 1,
    max_month = 3,
    min_lat   = 40,
    max_lat   = 60,
    min_depth = 40,
    max_depth = 60,
    min_iso = 100,
    max_iso=2000,
    max_dist = 30,
    min_dist = 0

  ), #long extent shelf break
  yolk = list( # yolk
    min_month = 1,
    max_month = 4,
    min_lat   = 31,
    max_lat   = 36,
    min_depth = 50, 
    max_depth = 100, 
    min_iso = 100,
    max_iso=2000 
  ), #long extent shelf break
  JAc = list( # JAc
    min_month = 1,
    max_month = 4,
    min_lat   = 34.5,
    max_lat   = 42.5,
    min_depth = 30,
    max_depth = 130,
    max_dist = 30,
    min_dist = 0
  ),
    JAs = list( # JAs
    min_month = 1,
    max_month = 4,
    min_lat   = 31,
    max_lat   = 34.5,
    min_depth = 30,
    max_depth = 130,
    max_dist = 30,
    min_dist = 0
  ),
    MSc = list( # MSc
    min_month = 5,
    max_month = 9,
    min_lat   = 34.5,
    max_lat   = 42.5,
    min_depth = 30,
    max_depth = 130,
    max_dist = 30,
    min_dist = 0
  ),
    MSs = list( # MSs
    min_month = 5,
    max_month = 9,
    min_lat   = 31,
    max_lat   = 34.5,
    min_depth = 30,
    max_depth = 130,
    max_dist = 30,
    min_dist = 0
  ),
    MSn = list( # MSn
    min_month = 5,
    max_month = 9,
    min_lat   = 42.5,
    max_lat   = 47,
    min_depth = 30,
    max_depth = 130,
    max_dist = 30,
    min_dist = 0
  ),
    larv = list( # larv
    min_month = 2,
    max_month = 5,
    min_lat   = 31,
    max_lat   = 36,
    min_depth = 50,
    max_depth = 200,
    min_iso = 100,
    max_iso=2000 
  ),
  latelarv = list( # latelarv
    min_month = 3,
    max_month = 6,
    min_lat   = 31,
    max_lat   = 37,
    min_depth = 50,
    max_depth = 300,
    min_iso = 100,
    max_iso=2000 
  ),
  late = list( # late
    min_month = 3,
    max_month = 6,
    min_lat   = 33.5,
    max_lat   = 34.5,
    min_depth = 75,
    max_depth = 275
  )
)

```

Figuring out why the distance variable wouldn't map onto the MOM6 data was a nightmare

```{r echo=FALSE}

# 1. Load local NetCDF datasets using xarray
# depth is used to filter longitudinal extent based on isobath
ds_depth <- xr$open_dataset("../STATIC/ocean_static.deptho.nc")
# depth is used to filter longitudinal extent based on distance from shore
# 1. Open the raw file
ds_dist_raw <- xr$open_dataset("../STATIC/mom6-nep_regrid_dist2coast_CCS.nc")

# 2. Safely isolate the true geographic decimal vectors from the 2D matrix arrays
# Since they are stored as variables, we can extract the true linear coordinates
true_lat_vector <- py_to_r(ds_dist_raw[["lat"]]$values[, 1])
true_lon_vector_raw <- py_to_r(ds_dist_raw[["lon"]]$values[1, ])

# 3. Convert Longitude map from (-180 to 180) into (0 to 360) global scale
# This ensures that -125 degrees west maps to 235 degrees global
true_lon_vector <- (true_lon_vector_raw %% 360)

# 4. Wipe out the old placeholder data variables to prevent name collisions
ds_dist_clean <- ds_dist_raw$drop_vars(list("lat", "lon"))

# 5. Overwrite the generic index coordinates with your true geographical decimals
ds_dist_assigned <- ds_dist_clean$assign_coords(list(
  latitude = true_lat_vector,
  longitude = true_lon_vector
))

# 6. Standardize the dimension labels to match your main data exactly
ds_dist <- ds_dist_assigned$rename(list(latitude = "lat", longitude = "lon"))

# Verify the changes (You will see active asterisks * next to true decimal coordinates!)
print(ds_dist)
```

```{r echo=FALSE}

# 2. Connect to Earthmover Arraylake
client <- arraylake$Client()
repo <- client$get_repo("NOAA-PMEL/cefi-nep-hindcast-monthly")
session <- repo$readonly_session(branch = "main")

# Load aux group Zarr data - this is data that is not a direct main output (it is calculated from model output)
aux <- xr$open_zarr(session$store, zarr_format = 3L, group = "regrid/aux1")
# Load the main group Zarr data 
main <- xr$open_zarr(session$store, zarr_format = 3L, group = "regrid/main")
# Use list() to match Python's double bracket [['MLD_003']] dataset subsetting
ds_mld <- aux[list("MLD_003")]
ds_temp = main[list('thetao')]
ds_xv = main[list('uo_rotate')]
ds_yv = main[list('vo_rotate')]
ds_ssh = aux[list('ssh')]

```

```{r echo=FALSE}
print_ds_info <- function(dataset, var_name, nan_bool = FALSE) {
  
  # Access the variable array using standard R list indexing
  var_array <- dataset[[var_name]]
  
  # In Python, .size and .shape are attributes
  denominator <- py_to_r(var_array$size)
  shape <- py_to_r(var_array$shape)
  
  # Format shape into a clean string (e.g., "(12, 360, 720)")
  shape_str <- paste0("(", paste(shape, collapse = ", "), ")")
  
  cat(sprintf("%d records with dimensions %s\n", denominator, shape_str))
  
  if (nan_bool) {
    # Call NumPy's isfinite on the python data object
    # .sum() converts the boolean array to a count
    numerator <- py_to_r(np$isfinite(var_array$data)$sum())
    percent <- (numerator / denominator) * 100
    
    cat(sprintf("%d records (%.2f%%) are not np.nan\n", numerator, percent))
  }
  
  return(invisible(NULL))
}
```


Developing a function for 2-dimensional data. This is relevant for MLD, SSH, and SST.
```{r echo=FALSE}
py <- import_builtins()
annual_mean_2d <- function(var_dataset, long_dataset, ds_var_name, df_var_name, filter_dict,stage, precond, print_counts = TRUE) {
  
  # Start timer
  start_time <- Sys.time()
  cat(paste0("---", df_var_name, "---\n"))
  
  # Print initial info
  print_ds_info(var_dataset, ds_var_name, print_counts)
  
  # Merge variable with mask file
  ds_merged <- xr$merge(list(var_dataset, long_dataset))
  print_ds_info(ds_merged, ds_var_name, print_counts)
  
  # Create list with months (R vectors auto-convert to Python lists)
  month_list <- seq(filter_dict[[stage]][["min_month"]], filter_dict[[stage]][["max_month"]])
  
  # Select times and latitudes
  # Using xr$slice for slicing dimensions via reticulate
  ds_selected <- ds_merged$sel(
    time = ds_merged$time$dt$month$isin(month_list), 
    lat = py$slice(filter_dict[[stage]][["min_lat"]], filter_dict[[stage]][["max_lat"]])
  )
  print_ds_info(ds_selected, ds_var_name, print_counts)
  
  # Mask longs using bitwise AND (&) on the xarray DataArrays
  ds_masked <- ds_selected$where(
    (ds_selected$deptho >= filter_dict[[stage]][['min_iso']]) & (ds_selected$deptho <= filter_dict[[stage]][['max_iso']])
  )
  print_ds_info(ds_masked, ds_var_name, print_counts)
  
  # Calculate mean grouped by year
  # Dim parameter accepts an R vector which automatically passes as a Python list
  ds_mean <- ds_masked[[ds_var_name]]$groupby('time.year')$mean(dim = c('time', 'lon', 'lat'))
  
  # Create pandas dataframe and automatically convert it to an R data.frame
  # Setting convert = TRUE ensures it returns a native R data frame
  ds_df <- py_to_r(ds_mean$to_dataframe())
  
    # --- ADDED PRECOND LAG LOGIC ---
  # If precond is TRUE, add a 1-year lag to the timeline data index
  if (precond) {
    cat("Applying 1-year environmental precondition lag (Years + 1)...\n")
    # In a default pandas to_dataframe() conversion with groupby('time.year'),
    # the years are stored as the index (rownames) of the R data frame.
    current_years <- as.numeric(rownames(ds_df))
    lagged_years <- current_years + 1
    rownames(ds_df) <- as.character(lagged_years)
  }
  
  # Rename column from dataset to dataframe variable name in R syntax
  names(ds_df)[names(ds_df) == ds_var_name] <- df_var_name
  
  # End timer and print runtime
  end_time <- Sys.time()
  runtime_minutes <- as.numeric(difftime(end_time, start_time, units = "mins"))
  cat(sprintf("Runtime: %.2f minutes\n", runtime_minutes))
  
  # Write output to CSV using standard R engine
  #write.csv(ds_df, paste0("Output/hake_MOM6_", df_var_name, ".csv"), row.names = TRUE)
  
  return(ds_df)
}

```

Is this function the speediest? No. But it is set up to be extremely efficient in how the data is loaded to minimize ram. With how it is set up it should use less than 1GB of ram and each individual variable should take less than 10 seconds to run. If you are running 50 variables it will take < 10 minutes to run and you won't crash R (my alternative formulation ran in 1 minute but if you had less than 120 gb of ram...good luck!). Ideally, this will make the code reproducible to daily data
```{r echo=FALSE}
annual_mean_2d_v2 <- function(var_dataset, long_dataset, distance, ds_var_name, df_var_name, filter_dict, stage, precond, print_counts = TRUE) {
  start_time <- Sys.time()
  cat(paste0("---", df_var_name, "---\n"))
  
  # Fetch native Python built-ins for slicing syntax
  py <- import_builtins()
  
  # 1. Temporal filter on the main array (1D operation)
  month_list <- seq(filter_dict[[stage]][["min_month"]], filter_dict[[stage]][["max_month"]])
  var_selected <- var_dataset$sel(time = var_dataset$time$dt$month$isin(month_list))
  
  # Extract year timelines safely
  py_years <- var_selected$time$dt$year$values
  r_years <- py_to_r(py_years)
  
  # Extract geographical coordinates cleanly as floating decimals
  min_lat <- as.numeric(filter_dict[[stage]][["min_lat"]])
  max_lat <- as.numeric(filter_dict[[stage]][["max_lat"]])
  
  # 2. Slice spatial coordinate boundaries on the main 1D variable array FIRST
  # This isolates only your 20-latitude window right out of the box
  var_selected <- var_selected$sel(lat = py$slice(min_lat, max_lat))
  var_array <- var_selected[[ds_var_name]]
  
  # 3. Dynamic Coordinate Name Standardization for long_dataset
  cat("Standardizing mask coordinate names...\n")
  dim_keys <- py_to_r(py$list(long_dataset$dims))
  
  rename_list <- list()
  if ("latitude" %in% dim_keys)  rename_list$latitude <- "lat"
  if ("longitude" %in% dim_keys) rename_list$longitude <- "lon"
  
  if (length(rename_list) > 0) {
    long_dataset <- long_dataset$rename(rename_list)
  }
  
  # 4. Use Xarray's interp_like to ensure alignment
  cat("Interpolating mask grid onto variable layout...\n")
  if (distance) {
    mask_sliced <- long_dataset[["dist"]]$interp_like(var_array, method = "nearest")
    py_mask <- (mask_sliced$lat >= min_lat) & 
               (mask_sliced$lat <= max_lat) & 
               (mask_sliced >= filter_dict[[stage]][['min_dist']]) & 
               (mask_sliced <= filter_dict[[stage]][['max_dist']]) & 
               (!mask_sliced$isnull())
  } else {
    mask_sliced <- long_dataset[["deptho"]]$interp_like(var_array, method = "nearest")
    py_mask <- (mask_sliced$lat >= min_lat) & 
               (mask_sliced$lat <= max_lat) & 
               (mask_sliced >= filter_dict[[stage]][['min_iso']]) & 
               (mask_sliced <= filter_dict[[stage]][['max_iso']]) & 
               (!mask_sliced$isnull())
  }
  
  # 5. Apply the mask layer
  cat("Applying mask layer...\n")
  var_masked <- var_array$where(py_mask)
  
  # Identify remaining spatial dimensions to collapse
  all_dims <- var_masked$dims
  spatial_dims_to_drop <- all_dims[all_dims != "time"]
  
  # 6. Collapse spatial dimensions natively inside Python
  cat("Collapsing spatial dimensions natively...\n")
  spatial_mean_series <- var_masked$mean(dim = spatial_dims_to_drop, skipna = TRUE)
  
  # 7. Pull the aggregated numeric timelines into R environment
  cat("Extracting values safe from memory overhead...\n")
  raw_values <- py_to_r(spatial_mean_series$values)
  
  
  
  # Manually construct a lightweight R data frame
  temp_df <- data.frame(
    year = r_years,
    value = raw_values
  )
  
  # 8. Aggregate annual timeline averages using R's engine
  cat("Aggregating annual timeline numbers in R...\n")
  ds_df <- aggregate(value ~ year, data = temp_df, FUN = mean, na.rm = TRUE)
  
      # --- ADDED PRECONDITIONING LAG LOGIC ---
  # If preconditioning is TRUE, add a 1-year lag to the timeline data index
  if (precond) {
    cat("Applying 1-year environmental precondition lag (Years + 1)...\n")
    # In a default pandas to_dataframe() conversion with groupby('time.year'),
    # the years are stored as the index (rownames) of the R data frame.
    current_years <- as.numeric(rownames(ds_df))
    lagged_years <- current_years + 1
    rownames(ds_df) <- as.character(lagged_years)
  }
  # Format column and index headers
  rownames(ds_df) <- ds_df$year
  ds_df$year <- NULL
  names(ds_df) <- df_var_name
  
  # Explicitly trigger system garbage collection
  rm(temp_df, raw_values, r_years)
  gc()
  
  # Print script runtime metrics
  end_time <- Sys.time()
  runtime_minutes <- as.numeric(difftime(end_time, start_time, units = "mins"))
  cat(paste0("Runtime: ", round(runtime_minutes, 2), " minutes\n"))
  
  return(ds_df)
}

```

Running functions for the 2d variables
```{r echo=FALSE}
#Run MLD code
MLDegg = annual_mean_2d_v2(ds_mld, ds_depth,FALSE, "MLD_003", "MLDegg", lifestage_dict, 'egg',precond=FALSE, TRUE)  # 1993 should = 47.29985 check if correct...

MLDlatelarv = annual_mean_2d_v2(ds_mld, ds_depth,FALSE, "MLD_003", "MLDlatelarv", lifestage_dict, 'latelarv', precond=FALSE, TRUE)

MLDyolk = annual_mean_2d_v2(ds_mld, ds_depth, FALSE,"MLD_003", "MLDlateyolk", lifestage_dict, 'yolk',precond=FALSE, TRUE)

MLDlarv = annual_mean_2d_v2(ds_mld, ds_depth, FALSE,"MLD_003", "MLDlateyolk", lifestage_dict, 'yolk',precond=FALSE,TRUE)

#Run SSH code
SSHjac = annual_mean_2d_v2(ds_ssh, ds_dist, TRUE,"ssh", "SSHjac", lifestage_dict, 'JAc',precond=FALSE,TRUE)

SSHjas = annual_mean_2d_v2(ds_ssh, ds_dist, TRUE,"ssh", "SSHjas", lifestage_dict, 'JAs',precond=FALSE,TRUE)

SSHmsc = annual_mean_2d_v2(ds_ssh, ds_dist, TRUE,"ssh", "SSHmsc", lifestage_dict, 'MSc',precond=FALSE,TRUE)

SSHmss = annual_mean_2d_v2(ds_ssh, ds_dist, TRUE,"ssh", "SSHmss", lifestage_dict, 'MSs',precond=FALSE,TRUE)

```